---
title: "Chimeric karyotypes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chimeric karyotypes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(karyoparser)
```

A `//` in an ISCN karyotype string separates two independent cell
populations -- typically host tissue (before `//`) and a donor/engrafted
population (after `//`), as seen post-transplant. karyoparser only parses one
clone per row, so it has to pick a side; the `on_chimeric` argument (on
`parse_karyo()` and `preprocess_karyo()`) controls which one.

## `on_chimeric`: choosing a clone

- `"default"`: host clone (before `//`) for normal chimeras; donor clone
  (after `//`) for zero-host chimeras (a leading `.//` prefix -- see below).
- `"host"`: host clone only; zero-host chimeras return `NA`.
- `"donor"`: everything after `//` (the donor population).

```{r host-donor}
parse_karyo(
  "47,XY,+8[15]//46,XX[5]",
  on_issues = "warn",
  on_chimeric = "host"
) |>
  dplyr::select(preprocessed_karyotype, chimeric_clone, tris8)

parse_karyo(
  "46,XX[15]//47,XY,+8[5]",
  on_issues = "warn",
  on_chimeric = "donor"
) |>
  dplyr::select(preprocessed_karyotype, chimeric_clone, tris8)
```

`chimeric_karyotype` is `1` for both rows (any `//` separator), and
`chimeric_clone` records which side was actually parsed.

## Zero-host chimeras

A leading `.//` (or `//`) prefix marks a **zero-host chimera**: a donor-only
population with no host metaphases. Under `"default"` or `"donor"` the donor
clone is parsed as usual; under `"host"` there is no host clone to parse, so
the row is `NA` -- but it's a *policy*-induced `NA`, not a data-quality
problem, so the row still counts as fixable, not unfixable:

```{r zero-host}
ck <- check_karyo(".//46,XX,+8[10]")
ck |>
  dplyr::select(karyotype, fixable, unfixable, zero_host_chimera)

parse_karyo(".//46,XX,+8[10]", on_issues = "warn", on_chimeric = "host") |>
  dplyr::select(preprocessed_karyotype, chimeric_clone)
parse_karyo(".//46,XX,+8[10]", on_issues = "warn", on_chimeric = "default") |>
  dplyr::select(preprocessed_karyotype, chimeric_clone, tris8)
```

## Multiple separators are always unfixable

A row with two or more `//` separators (`multiple_chimeric_separator`) has no
well-defined host/donor split, so it's always unfixable and returns `NA`
regardless of `on_chimeric` or `on_issues`:

```{r multiple-separator}
check_karyo("46,XX//47//48") |>
  dplyr::select(karyotype, fixable, unfixable)
parse_karyo("46,XX//47//48", on_issues = "preprocess", verbose = FALSE) |>
  dplyr::select(original_karyotype, preprocessed_karyotype, unfixable_error)
```

## `on_issues` is orthogonal to chimeric handling

Chimeric rows are always clone-selected and parsed, in every `on_issues`
mode -- a chimeric row only becomes `NA` when its selected clone is itself
unparseable (or the row has multiple separators). A non-chimeric issue
elsewhere in the batch still triggers `on_issues = "stop"`'s error; chimeric
rows never trigger it themselves:

```{r stop-with-chimera, error = TRUE}
parse_karyo(c("46,XX[15]//47,XY,+8[5]", ".46,XX,del(5)(q13)[10]"))
```

```{r stop-chimera-only}
parse_karyo("46,XX[15]//47,XY,+8[5]", verbose = FALSE) |>
  dplyr::select(preprocessed_karyotype, chimeric_clone)
```

The second call succeeds under the default `on_issues = "stop"` because the
only issue present is the chimeric separator itself.

## Clone selection and `karyo_preprocessed`

`preprocess_karyo()` bakes its `on_chimeric` choice into the `preprocessed`
column. When that `karyo_preprocessed` tibble is passed to `parse_karyo()`,
the same choice is inherited automatically -- passing an explicit,
conflicting `on_chimeric` at parse time raises an error, since the donor/host
split can't be un-done after the fact:

```{r preprocessed-inherit}
pp <- preprocess_karyo("46,XX[15]//47,XY,+8[5]", on_chimeric = "donor")
parse_karyo(pp) |>
  dplyr::select(preprocessed_karyotype, chimeric_clone, tris8)
```

```{r preprocessed-conflict, error = TRUE}
parse_karyo(pp, on_chimeric = "host")
```

Re-run `preprocess_karyo(..., on_chimeric = "host")`, or pass the raw
karyotypes straight to `parse_karyo(..., on_chimeric = "host")` instead.
