---
title: "Writing custom rule tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Writing custom rule tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(karyoparser)
```

`myeloid_rules` is `parse_karyo()`'s default rule set, but it's just a
`karyo_rules` table -- a two-column schema you can build yourself with
`validate_rules()` for a different disease area, panel, or one-off lesion.

## The `karyo_rules` schema

A rules table needs exactly two columns:

- `flag_name`: a unique label per rule. Can use ISCN-style punctuation for
  readability (e.g. `"t(9;22)(q34;q11)"`) -- the output column is a sanitized
  version, with runs of non-alphanumeric characters replaced by `_`.
- `regex`: the pattern matched against each token. Rules fire independently
  -- there's no priority ranking between them, so if two patterns can match
  the same flag, combine them into one `regex` with `|` rather than using two
  rows.

```{r validate}
my_rules <- validate_rules(tibble::tibble(
  flag_name = "t(X;18)(p11;q11)",
  regex = "t\\(X;18\\)\\(p11;q11\\)|t\\(18;X\\)\\(q11;p11\\)"
))
my_rules

result <- parse_karyo(
  "46,XX,t(X;18)(p11;q11)",
  rules = my_rules,
  on_issues = "warn",
  verbose = FALSE
)
result |>
  dplyr::select(t_X_18_p11_q11, general_translocation)
```

The output column (`t_X_18_p11_q11`) is free of characters (`(`, `)`, `;`)
that would otherwise force backtick-quoting (`` r$`t(X;18)(p11;q11)` ``) in
interactive use.

## Uniqueness is checked two ways

`validate_rules()` rejects duplicate `flag_name` values -- each one maps to
exactly one output column:

```{r dup-flag-name, error = TRUE}
validate_rules(tibble::tibble(
  flag_name = c("del(5q)", "del(5q)"),
  regex = c("del\\(5q", "del\\(5\\)\\(q")
))
```

It also rejects distinct `flag_name` values that collide once *sanitized*
into the same column name, even though the raw names differ:

```{r sanitize-collision, error = TRUE}
validate_rules(tibble::tibble(
  flag_name = c("t(9;22)", "t(9,22)"),
  regex = c("t\\(9;22\\)", "t\\(9;22\\)")
))
```

## Combining rule tables

`parse_karyo()`'s `rules` argument also accepts a `list()` of `karyo_rules`
tables, combining their flags:

```{r combine}
lesion_a <- validate_rules(tibble::tibble(
  flag_name = "flagA",
  regex = "t\\(9;22\\)\\(q34;q11\\)"
))
lesion_b <- validate_rules(tibble::tibble(
  flag_name = "flagB",
  regex = "del\\(5\\)\\(q13q33\\)"
))

result <- parse_karyo(
  c("46,XX,t(9;22)(q34;q11)", "46,XX,del(5)(q13q33)"),
  rules = list(myeloid_rules, lesion_a, lesion_b),
  on_issues = "warn",
  verbose = FALSE
)
result |>
  dplyr::select(t_9_22_q34_q11, flagA, flagB)
```

`flag_name` must stay unique *across* the combined tables too -- there's no
cross-table priority to resolve a clash, so combining two tables that both
define, say, `"shared_flag"` errors instead of silently keeping one:

```{r combine-clash, error = TRUE}
set_a <- validate_rules(tibble::tibble(
  flag_name = "shared_flag",
  regex = "t\\(9;22\\)\\(q34;q11\\)"
))
set_b <- validate_rules(tibble::tibble(
  flag_name = "shared_flag",
  regex = "del\\(5\\)\\(q13q33\\)"
))
parse_karyo(
  "46,XX,t(9;22)(q34;q11)",
  rules = list(set_a, set_b),
  on_issues = "warn",
  verbose = FALSE
)
```

## Rules vs. general flags

General structural categories -- translocations, deletions, inversions,
additions, dicentrics, isodicentrics, pseudodicentrics, isochromosomes,
rings, insertions, duplications, triplications, markers, double minutes,
derivatives -- are
already detected universally by the parser as `general_*` columns,
independently of whatever `rules` table you pass. You don't need a rule for
"any translocation" -- write rules only for the *specific* lesions your
analysis cares about.

## Restricting output with `columns`

The full output is wide -- 170 columns for the default `myeloid_rules`.
Pass `columns` to `parse_karyo()` to keep only the groups you need; metadata
and status columns (`original_karyotype`, `fixable_error`,
`chimeric_karyotype`, ...) are always included regardless. Valid values:
`"classification"`, `"rule"`, `"general"`, `"aneuploidy"`, `"summary"`,
`"der_loss"` (see `vignette("karyoparser")` for what each group
contains).

```{r columns}
parse_karyo(
  c("46,XX", "47,XY,+21,t(9;22)(q34;q11)"),
  on_issues = "warn",
  verbose = FALSE,
  columns = c("classification", "general")
)
```

`columns = "rule"` reflects whichever rules table you passed -- including a
custom one:

```{r columns-custom-rules}
parse_karyo(
  "46,XX,t(X;18)(p11;q11)",
  rules = my_rules,
  on_issues = "warn",
  verbose = FALSE,
  columns = "rule"
)
```
