---
title: "Cleaning dirty karyotype strings"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cleaning dirty karyotype strings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(karyoparser)
```

Real-world karyotype strings are messy: stray whitespace, HTML entities,
missing commas, FISH annotations tacked onto the end, and outright
out-of-scope constructs like mosaicism. This vignette covers `check_karyo()`,
`preprocess_karyo()`, and the `on_issues` argument to `parse_karyo()` in
detail.

## Fixable vs. unfixable

Every issue `check_karyo()` finds is either **fixable** (a formatting
artifact) or **unfixable** (a structural problem). `preprocess_karyo()` only
touches the fixable kind -- unfixable rows always come out `NA`, no matter
which `on_issues` mode you use.

```{r issue-catalog}
ck <- check_karyo(example_karyotypes)
names(ck)
```

Columns after `karyotype`/`fixable`/`unfixable` are one 0/1 flag per issue
type -- unfixable types first (alphabetical), then fixable types
(alphabetical):

- **Fixable**: `chimeric_separator`, `count_sex_separator` (e.g. `46XY` or
  `45.XY` missing the comma before the sex complement), `embedded_newline`,
  `fish_notation`, `html_entities`, `leading_dot`, `mar_space`,
  `midstring_linewrap`, `missing_sex_comma` (e.g. `46,XX der(...)` missing the
  comma after the sex complement), `non_clonal_sca` (an `ncSCA` token,
  standalone or parenthesized), `trailing_narrative`, `unicode_notation`,
  `zero_host_chimera`.
- **Unfixable**: `constitutional_sex_complement`, `empty`, `invalid_idem`,
  `mosaic_karyotype`, `multiple_chimeric_separator`, `no_chromosome_count`,
  `no_sex_complement`, `unbalanced_brackets`, `unbalanced_parentheses`,
  `unparseable_bracket`, `updated_iscn`.

## Out-of-scope constructs take precedence

Mosaicism (`mos`) and constitutional abnormalities (a sex complement with a
`c` suffix, e.g. `47,XXYc`) are deliberately out of scope and flagged
unfixable -- even though the string may otherwise look structurally complete.
These take precedence over `no_chromosome_count`/`no_sex_complement`, so a
present count or complement isn't mislabeled as missing:

```{r precedence}
ck <- check_karyo(c(
  "mos 47,XXY[10]/46,XY[5]",
  "47,XXYc[20]"
))
ck |>
  dplyr::select(
    karyotype,
    mosaic_karyotype,
    constitutional_sex_complement,
    no_chromosome_count,
    no_sex_complement
  )
```

Each row has a chromosome count and sex complement present, but
`no_chromosome_count`/`no_sex_complement` stay `0` -- the out-of-scope flag
fires instead.

## `ncSCA` tokens are silently stripped

Non-clonal single-cell abnormalities (an `ncSCA` token, standalone or
parenthesized) carry no structural detail of their own, so `preprocess_karyo()`
strips the token and parses whatever clone remains:

```{r ncsca}
preprocess_karyo(c("ncSCA[4]/46,XY[11]", "46,XX(ncSCA)[1]//46,XY[19]"))
```

## `preprocess_karyo()`'s cleaning pipeline

`preprocess_karyo()` is the only function that rewrites a string. It applies
fixes in a fixed order, then a final `normalize_iscn()` pass:

1. Trim leading/trailing whitespace
2. Replace unicode spaces (NBSP), dashes (em/en-dash), fullwidth characters
3. Collapse embedded newlines/tabs to spaces
4. Decode HTML entities (`&lt;` -> `<`, `&gt;` -> `>`, `&amp;` -> `&`)
5. Strip leading dot(s) before a digit (e.g. `.46,XX` -> `46,XX`)
6. Strip FISH/`nuc ish` annotation (suffix or mid-clone before metaphase count)
7. Strip trailing narrative: `] .text` -> `]`; `) Capital text` -> `)`
8. Collapse mid-string line-wrap artifacts (`, .der(...)` -> `,der(...)`)
9. Insert missing comma after sex chromosome complement
10. Remove space between count and `mar` token (`+1~4 mar` -> `+1~4mar`)
11. `normalize_iscn()`: whitespace collapsing, delimiter tightening,
    idem/sl/cp normalization

```{r preprocess}
messy <- c(
  ".46,XX,del(5)(q13)[10]",
  "46,XX,t(9;22)(q34;q11)[15] &lt;AML&gt;",
  "46,XX[cp 20]"
)
preprocess_karyo(messy, verbose = TRUE)
```

`preprocess_karyo()` accepts a `karyo_check` object directly (as returned by
`check_karyo()`), reusing its cached assessment instead of re-scanning:

```{r reuse-check}
ck <- check_karyo(example_karyotypes)
pp <- preprocess_karyo(ck)
pp
```

## `on_issues`: how `parse_karyo()` handles issues

`parse_karyo()`'s `on_issues` argument controls what happens to dirty and
structural rows. It has no effect on chimeric handling -- see
`vignette("chimeric-karyotypes")` -- chimeric rows are always clone-selected
and parsed regardless of `on_issues`; only a non-chimeric issue on the *same*
row prevents that.

### `"stop"` (default)

Raises an error immediately if any non-chimeric issue is found, listing
fixable/unfixable issue types and counts:

```{r stop-mode, error = TRUE}
parse_karyo(c("46,XX", ".46,XX,del(5)(q13)[10]", "mos 47,XXY[10]"))
```

A non-chimeric issue anywhere in the batch triggers this error even if other
rows in the same batch are chimeric -- chimeric rows themselves never trigger
it.

### `"preprocess"`

Applies `preprocess_karyo()` to dirty rows; rows still broken afterwards
(the unfixable kind) come back `NA`:

```{r preprocess-mode}
parse_karyo(
  c("46,XX", ".46,XX,del(5)(q13)[10]", "mos 47,XXY[10]"),
  on_issues = "preprocess",
  verbose = TRUE
)
```

### `"warn"`

Returns `NA` for all dirty/structural issue rows and emits a warning -- no
fixing is attempted:

```{r warn-mode, warning = TRUE}
parse_karyo(
  c("46,XX", ".46,XX,del(5)(q13)[10]", "mos 47,XXY[10]"),
  on_issues = "warn"
)
```

## Next steps

- `vignette("chimeric-karyotypes")` -- clone selection for `//`-separated
  chimeric karyotypes.
- `vignette("custom-rules")` -- writing your own rule tables.
