---
title: "Getting Started with karyoparser"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with karyoparser}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(karyoparser)
```

karyoparser turns ISCN karyotype strings into a wide tibble of binary
features. This vignette walks through the three pipeline functions, the
`example_karyotypes` dataset that ships with the package, and how to read the
output.

## The three pipeline functions

| Function / Object | Description |
|---|---|
| `parse_karyo()` | Parse karyotype strings into a wide feature tibble |
| `check_karyo()` | Scan for formatting artifacts and structural errors |
| `preprocess_karyo()` | Clean and normalize dirty strings |
| `validate_rules()` | Build a custom `karyo_rules` object |
| `myeloid_rules` | Default rule set for myeloid neoplasms (data object) |
| `example_karyotypes` | Synthetic karyotypes for trying out the package (data object) |

`check_karyo()`, `preprocess_karyo()`, and `parse_karyo()` each do one job,
and none of them need the others -- call whichever one you actually need. The
first time you're working with a new dataset, though, run all three in
order -- diagnose, clean, extract -- so you actually see what's wrong with the
data before it quietly turns into missing rows.

- **`check_karyo()` -- diagnose.** Read-only. Scans the input and returns a
  `karyo_check` tibble flagging issue types (dirty markers, chimeric
  separators, structural errors), plus `fixable`/`unfixable` summary columns.
- **`preprocess_karyo()` -- clean.** The only function that rewrites a string.
  Fixes what's fixable and returns a narrow `karyo_preprocessed` tibble
  (`original`, `preprocessed`, `status`).
- **`parse_karyo()` -- extract.** Feature extraction only -- no cleaning of
  its own. `on_issues = "preprocess"` can invoke the same check/fix logic
  internally, so a single call is often enough; chain the three explicitly
  when you also want the diagnostic or cleaned tibble in its own right. See
  `vignette("data-cleaning")` for the full `on_issues` reference.

## Chaining pipeline steps

`check_karyo()` and `preprocess_karyo()` tag their output with an extra S3
class, so you can pipe straight into the next step without re-specifying
arguments:

| Function | Class | Notes |
|---|---|---|
| `check_karyo()` | `karyo_check` (+ `tbl_df`) | Accepted as `preprocess_karyo()` input; the assessment is reused instead of re-run. |
| `preprocess_karyo()` | `karyo_preprocessed` (+ `tbl_df`) | Accepted as `parse_karyo()` input; the `preprocessed` column and cached id/issue info are reused automatically. |
| `parse_karyo()` | plain `tbl_df` | No custom class. |

`parse_karyo()`'s output also carries a `karyoparser_version` attribute
(`attr(result, "karyoparser_version")`) set to the installed package version,
so you can trace a saved result back to whatever version produced it.

## A first pass, with `example_karyotypes`

`example_karyotypes` ships with the package -- 100 synthetic ISCN strings
covering one example per `myeloid_rules` flag, general structural
aberrations, aneuploidy-only cases, balanced/unbalanced translocations,
complex and monosomal karyotypes, chimeric/multi-clone cases, and a batch of
messy or out-of-scope strings.

```{r example-data}
example_karyotypes

ck <- check_karyo(example_karyotypes)
table(fixable = ck$fixable, unfixable = ck$unfixable)
```

Fixable rows are formatting artifacts (stray whitespace, HTML entities, a
missing comma, ...); unfixable rows are structural problems (no chromosome
count, mosaicism, ...) that always come out `NA`. `preprocess_karyo()` only
touches the fixable kind. See `vignette("data-cleaning")` for the full
breakdown of issue types and how `on_issues` handles each.

```{r pipeline}
pp <- preprocess_karyo(example_karyotypes)
result <- parse_karyo(pp, on_issues = "preprocess")
result[, c("sample_id", "chromosome_count", "general_translocation", "complex_karyotype")]
```

`parse_karyo(example_karyotypes, on_issues = "preprocess")` does the same
check-and-fix work in one call, without needing the intermediate
`karyo_check`/`karyo_preprocessed` tibbles.

## Data frame input

`parse_karyo()` also takes a data frame directly. It auto-detects the
karyotype column (common names like `karyotype`, `iscn`, `karyo`, ...) and an
optional id column (`sample_id`, `patient_id`, `id`, `mrn`, ...).

```{r df-input}
df <- data.frame(
  sample_id = c("S1", "S2", "S3"),
  iscn = c("46,XX", "47,XY,+21[10]/46,XY[5]", "46,XX,t(9;22)(q34;q11)[20]")
)

result <- parse_karyo(df, verbose = FALSE)
result[, c("sample_id", "original_karyotype", "chromosome_count", "tris21")]
```

## Output reference

### Output columns

Each row is one input karyotype; columns fall into a handful of groups by
provenance. The full output is wide -- 167 columns for the default
`myeloid_rules` -- so pass `columns` to `parse_karyo()` to keep just the
groups you want (see `vignette("custom-rules")` for details). Metadata and
status columns are always included regardless.

| Group | Columns | Type | Description |
|---|---|---|---|
| Metadata | `original_karyotype`, `preprocessed_karyotype` | character | Raw input; cleaned/normalized string that was parsed (NA for unfixable rows) |
| Classification | `normal_karyotype` | integer 0/1 | 1 if `46,XX` or `46,XY` exactly |
| Classification | `complex_karyotype` | integer 0/1 | 1 if >= 3 distinct aberrations across clones |
| Classification | `monosomal_karyotype` | integer 0/1 | 1 if >= 2 autosomal monosomies, or >= 1 monosomy + >= 1 structural aberration |
| Rule flags | *(one per `myeloid_rules` entry)* | integer 0/1 | Specific lesions -- see Aberration flags below |
| General flags | `general_translocation`, `general_deletion`, `general_inversion`, `general_addition`, `general_dicentric`, `general_isodicentric`, `general_pseudodicentric`, `general_isochromosome`, `general_ring`, `general_insertion`, `general_duplication`, `general_triplication`, `general_marker`, `general_derivative`, `balanced_translocation`, `unbalanced_translocation` | integer 0/1 | Universal structural-aberration detections, incl. translocation balance |
| Aneuploidy | `mono1`--`mono22`, `monoX`, `monoY`; `tris1`--`tris22`, `trisX`, `trisY` | integer 0/1 | Whole-chromosome loss/gain from `-`/`+` tokens |
| Summary | `comma_count_aberrations` | integer | Aberration count (max across clones; idem-expanded) |
| Summary | `chromosome_count` | integer | Count from the most abnormal eligible clone |
| Summary | `total_metaphases` | integer | Sum of bracket counts; NA if no brackets |
| Derived loss | `unbal_partial_loss_<arm>` (one per chromosome arm), `unbal_partial_loss` | integer 0/1 | Partial arm loss implied by an unbalanced der translocation |
| Status | `fixable_error`, `unfixable_error` | integer 0/1 | Row had a fixable / unfixable issue (unfixable rows are all NA) |
| Status | `chimeric_karyotype` | integer 0/1 | 1 if input contained a `//` chimeric separator |
| Status | `chimeric_clone` | character | Which clone was parsed for a chimeric row (`"host"`, `"donor"`, or NA) -- see `vignette("chimeric-karyotypes")` |

### Aberration flags

Every flag is `0` or `1`, and **every flag fires independently** -- there's no
priority or competition between them, so one token can light up several flags
at once. They come from three sources:

- **Specific rule flags** -- one per `myeloid_rules` entry, fired when its
  regex matches a token (e.g. `t(9;22)(q34;q11)`, `del(5q)`).
- **General structural flags** (`general_*`) -- universal, disease-agnostic
  detections (any translocation, deletion, inversion, addition, dicentric,
  isodicentric, pseudodicentric, isochromosome, ring, insertion, duplication,
  triplication, marker, derivative). Always on, regardless of which rule set
  you pass.
- **Aneuploidy flags** (`mono*`/`tris*`) -- whole-chromosome loss/gain from
  `-`/`+` tokens.

Since flags don't compete, a token can trip **all** its matches at once.
`t(9;11)(p21;q23)` sets the specific `t_9_11_p21_q23`, the family flag
`t_v_11q23`, and `general_translocation`; `del(5q)` sets both `del_5q` and
`general_deletion`.

- **Family detectors** (`t_v_11q23`, `t_v_11p15`, `t_3q26_v`, `t_5q`,
  `t_12p`) deliberately co-fire with the more specific flags -- they mean "any
  rearrangement involving this region/arm".
- **`*_other` variants** (e.g. `t_9_22_other`) are the complement of the
  canonical breakpoints: they fire for the same chromosome pair at
  *non-canonical* breakpoints only -- the exclusion lives in the regex itself,
  so a canonical token sets just the specific flag.

```{r rules}
myeloid_rules
```

### The CBF-AML exception to `monosomal_karyotype`

`monosomal_karyotype` fires when a row has >= 2 autosomal monosomies, or >= 1
monosomy plus >= 1 structural aberration -- except for core-binding-factor
AML. Rows carrying `t(8;21)(q22;q22)`, `inv(16)(p13q22)`, or
`t(16;16)(p13;q22)` are never classified monosomal, regardless of any
co-firing monosomy or structural flag, per clinical guidelines:

```{r cbf-override}
result <- parse_karyo(
  "45,XX,-7,t(8;21)(q22;q22)",
  on_issues = "warn",
  verbose = FALSE
)
result[, c("t_8_21_q22_q22", "mono7", "general_translocation", "monosomal_karyotype")]
```

Without the override, `-7` plus the translocation's structural flag would
otherwise satisfy the monosomal criteria.

## Next steps

- `vignette("data-cleaning")` -- the `on_issues` modes, the fixable/unfixable
  issue catalog, and `preprocess_karyo()`'s cleaning pipeline in detail.
- `vignette("chimeric-karyotypes")` -- how `//`-separated chimeric karyotypes
  are handled.
- `vignette("custom-rules")` -- writing your own rule tables and restricting
  `parse_karyo()`'s output with `columns`.
