---
title: "Cytogenetic risk stratification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cytogenetic risk stratification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(karyoparser)
```

`assign_risk()` turns a `parse_karyo()` result into a cytogenetic risk
category, under either the IPSS-R cytogenetic risk groups or the ELN 2022
risk classification.

## Read this first

**Neither column is a risk score.** Both are the *cytogenetic component* of a
classification that needs more than a karyotype:

- **IPSS-R** stratifies on five components -- cytogenetics, marrow blast
  percentage, haemoglobin, platelet count, and absolute neutrophil count.
  `assign_risk()` computes the first. An `ipssr_cyto_risk` of `Good` is not
  an IPSS-R risk group; it is one of five inputs to one.
- **ELN 2022 is a genetic classification**, not a cytogenetic one. Assigning
  it properly requires *NPM1*, *FLT3*-ITD, *CEBPA* bZIP, *TP53*, *ASXL1*,
  *RUNX1* and others. A normal karyotype is intermediate by cytogenetics, but
  favorable with mutated *NPM1* and no *FLT3*-ITD, or adverse with mutated
  *TP53*. Molecular findings routinely move a case across categories.

The columns are named `*_cyto_*` for this reason. Treat them as a
cytogenetics-derived input to a risk assessment, never as the assessment.

## Quick start

Apply one scheme per call and chain for both:

```{r quickstart}
k <- c(
  "46,XX",
  "46,XX,t(8;21)(q22;q22)",
  "46,XX,del(5)(q13q33)",
  "45,XY,-7",
  "49,XX,+8,+13,+21"
)

parse_karyo(k) |>
  assign_risk("ipssr") |>
  assign_risk("eln2022") |>
  dplyr::select(
    original_karyotype,
    ipssr_cyto_risk,
    ipssr_cyto_score,
    eln2022_cyto_risk
  )
```

Note the last two rows. `46,XX,del(5)(q13q33)` is `Good` under IPSS-R and
`Adverse` under ELN; `49,XX,+8,+13,+21` is `Poor` under IPSS-R (three
abnormalities) but `Intermediate` under ELN (the hyperdiploid carve-out,
below). Neither is a bug -- see
[Where the two schemes disagree](#where-the-two-schemes-disagree).

Both risk columns are **ordered factors**, so they sort and compare directly:

```{r ordered}
r <- assign_risk(parse_karyo(c("46,XX", "45,XY,-7")), "ipssr")
r$ipssr_cyto_risk[2] > r$ipssr_cyto_risk[1]
```

`assign_risk()` appends to the tibble it is given and changes nothing else,
so the parsed feature columns stay available alongside the categories.

## Counting aberrations

Both schemes ask how many abnormalities a karyotype carries. Both use
`distinct_aberrations` -- distinct normalized aberration tokens pooled across
all clones, which is the same count `complex_karyotype` thresholds at 3. Using
one count for both means a row can never be flagged complex while the risk
logic counts two.

Polysomy collapses: `+8,+8` is tetrasomy 8, one abnormality, not two.

```{r counting}
parse_karyo(c("49,XY,+8,+8,+21", "49,XY,+8,+13,+21")) |>
  dplyr::select(original_karyotype, distinct_aberrations, complex_karyotype)
```

## IPSS-R cytogenetic risk groups

| Category | Score | Definition |
|---|---|---|
| Very Good | 0 | isolated `-Y`, isolated `del(11q)` |
| Good | 1 | normal, isolated `del(5q)`/`del(12p)`/`del(20q)`, double including `del(5q)` |
| Intermediate | 2 | `del(7q)`, `+8`, `+19`, `i(17q)`, any other single or double |
| Poor | 3 | `-7`, `inv(3)`/`t(3q)`/`del(3q)`, double including `-7`/`del(7q)`, 3 abnormalities |
| Very Poor | 4 | more than 3 abnormalities |

```{r ipssr}
ipssr_examples <- c(
  "45,X,-Y",
  "46,XX,del(5)(q13q33)",
  "47,XX,+8",
  "46,XX,inv(3)(q21q26)",
  "43,XX,-5,-7,-18,del(5)(q13)"
)

parse_karyo(ipssr_examples) |>
  assign_risk("ipssr") |>
  dplyr::select(
    original_karyotype,
    distinct_aberrations,
    ipssr_cyto_risk,
    ipssr_cyto_score
  )
```

### Where the published table is silent

Two resolutions are applied deliberately. Both are worth confirming against
the source before you rely on them.

**A double containing both `del(5q)` and `-7`/`del(7q)`** satisfies the Good
row ("double including del(5q)") and the Poor row ("double including
-7/del(7q)") simultaneously. Poor wins, so the more adverse lesion is not
masked by its partner:

```{r ipssr-tiebreak}
parse_karyo("44,XX,del(5)(q13q33),-7") |>
  assign_risk("ipssr") |>
  dplyr::select(original_karyotype, ipssr_cyto_risk)
```

**`t(3q)` is read as 3q26 rearrangement**, not any 3q breakpoint, since the
IPSS-R row reflects *MECOM*/EVI1 biology. `del(3q)` is matched at any 3q
breakpoint.

### Karyotypes IPSS-R cannot score

A karyotype with no scoreable aberration token that is nonetheless not normal
returns `NA` rather than a guess. The usual case is `45,X`, where the loss
rides on the sex complement and `-X` cannot be distinguished from `-Y` -- one
would be Very Good, the other Intermediate:

```{r ipssr-na}
parse_karyo("45,X") |>
  assign_risk("ipssr") |>
  dplyr::select(original_karyotype, distinct_aberrations, ipssr_cyto_risk)
```

Rows the parser could not score at all are `NA` in the same way.

## ELN 2022 cytogenetic risk

Evaluated in strict precedence order:

1. **Favorable** -- `t(8;21)`, `inv(16)`, `t(16;16)`. Class-defining, so they
   outrank every adverse criterion below.
2. **Intermediate, by carve-out** -- `t(9;11)`/*MLLT3::KMT2A*.
3. **Adverse** -- `t(6;9)`, `t(v;11q23)`, `t(9;22)`, `t(8;16)`,
   `inv(3)(q21q26)`/`t(3;3)(q21;q26)`, `t(3q26;v)`, `-5`/`del(5q)`, `-7`,
   `-17`/abn(17p), complex karyotype, monosomal karyotype.
4. **Intermediate** -- anything else, including a normal karyotype.

ELN 2022 has no numeric score, so this scheme adds only a category column.

```{r eln}
eln_examples <- c(
  "46,XX,inv(16)(p13q22)",
  "46,XX",
  "46,XX,t(9;22)(q34;q11)",
  "45,XX,-7",
  "46,XX,i(17)(q10)"
)

parse_karyo(eln_examples) |>
  assign_risk("eln2022") |>
  dplyr::select(original_karyotype, eln2022_cyto_risk)
```

### The KMT2A carve-out

`t(9;11)` is an 11q23 rearrangement, so it also matches the *adverse*
`t(v;11q23)`/*KMT2A*-rearranged row. ELN gives `t(9;11)` precedence, so the
order matters: check it before the adverse rows, or every `t(9;11)` comes out
adverse.

```{r kmt2a}
kmt2a <- c("46,XX,t(9;11)(p21;q23)", "46,XX,t(11;19)(q23;p13)")

parse_karyo(kmt2a) |>
  assign_risk("eln2022") |>
  dplyr::select(
    original_karyotype,
    t_9_11_p21_q23,
    t_v_11q23,
    eln2022_cyto_risk
  )
```

Both rows have `t_v_11q23 = 1`; only the second is adverse.

### Class-defining lesions outrank adverse criteria

A favorable lesion wins even alongside monosomies and a complex karyotype --
the same logic already applied to `monosomal_karyotype`, which is forced to 0
for CBF-AML:

```{r eln-favorable}
parse_karyo("43,XX,-5,-7,-18,t(8;21)(q22;q22)") |>
  assign_risk("eln2022") |>
  dplyr::select(
    original_karyotype,
    complex_karyotype,
    monosomal_karyotype,
    eln2022_cyto_risk
  )
```

### The hyperdiploid carve-out

ELN's complex definition excludes hyperdiploid karyotypes with three or more
trisomies (or polysomies) and no structural abnormality. Adding a structural
abnormality, or a monosomy, brings complex back:

```{r hyperdiploid}
hd <- c(
  "49,XX,+8,+13,+21",
  "52,XX,+1,+6,+8,+9,+14,+21",
  "49,XX,+8,+13,+21,del(20)(q11)",
  "48,XX,+8,+13,-18",
  "49,XX,+8,+13,+21,+mar"
)

parse_karyo(hd) |>
  assign_risk("eln2022") |>
  dplyr::select(original_karyotype, complex_karyotype, eln2022_cyto_risk)
```

Note `complex_karyotype` stays `1` throughout. **The carve-out is applied
inside the ELN path only**, because it is specific to ELN: IPSS-R's complex
rows carry no such exclusion, and folding it into the shared column would
silently corrupt IPSS-R scoring. A marker chromosome counts as a structural
abnormality here, which is why the last row is adverse.

## Where the two schemes disagree

They are built for different diseases and genuinely differ. Isolated `del(5q)`
is the clearest case -- Good under IPSS-R, adverse under ELN:

```{r disagree-example}
parse_karyo(c("46,XX,del(5)(q13q33)", "47,XX,+8", "46,XX,t(8;21)(q22;q22)")) |>
  assign_risk("ipssr") |>
  assign_risk("eln2022") |>
  dplyr::select(original_karyotype, ipssr_cyto_risk, eln2022_cyto_risk)
```

Across the bundled example dataset:

```{r disagree-table}
parse_karyo(example_karyotypes, on_issues = "preprocess", verbose = FALSE) |>
  assign_risk("ipssr") |>
  assign_risk("eln2022") |>
  dplyr::count(ipssr_cyto_risk, eln2022_cyto_risk)
```

Off-diagonal cells are the schemes disagreeing by design, not errors.

## Judgment calls and limits

Beyond the IPSS-R resolutions above, the ELN implementation makes two calls
worth knowing about:

- **abn(17p)** is read as `-17`, `del(17p)`, `add(17p)` or `i(17q)`.
  Including `i(17q)` reflects that it entails 17p loss.
- **Partial 17p loss implied by an unbalanced derivative is not counted.**
  `unbal_partial_loss_*` is documented as separate from `del()`/`mono*`, and
  folding it in here would breach that boundary. If you want those rows
  treated as abn(17p), combine `unbal_partial_loss_17p` yourself.

Two further limits apply to both schemes:

- **Custom rule tables.** Risk assignment reads specific [myeloid_rules]
  flags. Passing a different `rules` table, or narrowing `columns` so those
  flags are dropped, makes `assign_risk()` error rather than return a wrong
  category.
- **Multi-clone karyotypes** are scored on aberrations pooled across all
  clones, consistent with `complex_karyotype`. Neither published scheme
  prescribes a clone-selection policy.

## Next steps

- `vignette("karyoparser")` -- the full output column reference.
- `vignette("custom-rules")` -- writing rule tables and restricting output
  with `columns`.
